---
title: "Summarizing HTP Temporal Data"
author: "singhdj2@ksu.edu"
date: "August 8, 2019"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
The objective of today's exercise is to make you familiar with the processing of high-throughput phenotyping data. We will use plant height trait as an example and along the way will learn how to summarize and explore data. Some of the concepts/topics discussed here would be applicable to your research as well. 

# About the Data
The data was collected from two wheat experiments in South Asia in field season 2018 using an unmanned aerial vehicle a.k.a. drone. There were a total of 1200 wheat test plots (size 1.2m x 3m) in each experiment. The two experiments are identical in terms of genotypes and plot layout except these were planted at two different sowing dates (Optimum on 23-Oct, Late on 11 Nov). The main objective of this field experiment was to assess the effect of sowing dates on plant growth and yield. The data was collected at multiple times during the wheat growing season. Overall this dataset captures the dynamic changes in wheat height during its growth cycle. 

# Expected Outcome 
Through this exercise you will learn:

* How to process plot-level height data from multiple dates
* Model the growth dynamics of height through non-linear regression 
* Visualize the results and make sense of the trends 
* Compare the effects of sowing dates on wheat height and growth

## First we will load all the required packages
```{r load-packages, echo=FALSE}
#install.packages(pacman)  #run this line if pacman not installed already
library(pacman)
p_load(tidyverse,nlme,car,ggpubr,rcompanion) 
```

## First, let us create a logistic regression function to model wheat growth. This is just the definition of a function, we will call it during the later steps.  

## Quick question: Why is there any need to create functions? 

### Non-linear Logistic Regression Function
```{r model-definition}
runLogReg = function(dates, phenValue){
  # create column name strings for fit stats
  fit.stat.cols <- c("Min.max.accuracy","MAE","MAPE","MSE","RMSE",
                    "NRMSE.mean","NRMSE.median","NRMSE.mean.accuracy", "NRMSE.median.accuracy","Efron.r.squared","CV.prcnt")
  if(sum(!is.na(phenValue))==0){         # skip if no phenotype data
    pred = data.frame(x=NA,y=NA) %>%     # fill NA since no data
      nest()
    res <- data.frame(phi1=NA, phi2=NA, phi3=NA, 
                      flag="no-data", stringsAsFactors = F)
    fit.stats <- as.data.frame(matrix(nrow = 1,ncol = 11,
                                      dimnames = list(1,fit.stat.cols))) # empty fit.stats
    pred.res <- bind_cols(pred,res,fit.stats)      # bind columns
    return(pred.res)                     # return the data frame
    } 
  if(max(phenValue, na.rm=TRUE) < 60){   ## skip plots that never reach 60cm height
    pred = data.frame(x=NA,y=NA) %>%
      nest()
    res <- data.frame(phi1=NA, phi2=NA, phi3=NA, flag="short.max.ht", 
                      stringsAsFactors = F) # flag short plots
    fit.stats <- as.data.frame(matrix(nrow = 1,ncol = 11,
                                      dimnames = list(1,fit.stat.cols))) # empty fit.stats
    pred.res <- bind_cols(pred,res,fit.stats)
    return(pred.res)
  } 
  ## get Thermal Time
  days = dates 
  if(sum(!is.na(phenValue)) >= 9) {        # run only plots that have at least 9 time-points  
    # add some dummy variables for min and max height at 50-100 ThermalDays before/after the phenotyping range
    # buffer raw data to force start values to zero
    days = c(days, c(0,50,75,100), max(days,na.rm=T)+c(50,75,100,150))
    height = c(phenValue, c(0,0,0,0),rep(max(phenValue,na.rm=T),4))   #c(0,0,0,100,100,100))
    ##find initial starting values for phi2 and phi3,  fix phi1 at max pheno value
    phi1 = max(phenValue,na.rm=T)                      # maximum height parameter
    phi2 = coef(lm(logit(height/phi1)~days))[2]        # growth rate
    phi3 = -coef(lm(logit(height/phi1)~days))[1]/phi2  # time at half-max growth
    # Run logistic regression model
    growth_mod <- try(nls(height~ max(phenValue,na.rm=T)/(1+exp(-phi2*(days - phi3))), 
                          start=list(phi2=phi2,phi3=phi3), trace=FALSE), silent=TRUE)
    if(class(growth_mod)!="try-error"){              # this condition is to handle errors
      phi2 = coef(growth_mod)[1]
      phi3 = coef(growth_mod)[2]
      ## get predicted value
      pred = getPred(phi1, phi2, phi3) %>%
        nest()
      res = data.frame(phi1, phi2, phi3,flag="data.complete",
                       stringsAsFactors = F)         # combine all growth coeficients
      fit.stats <- accuracy(list(growth_mod),plotit = F)$Fit.criteria # gives pseudo-Rsq and other fit stats
      colnames(fit.stats) <- fit.stat.cols
      pred.res = bind_cols(pred,res,fit.stats)       # bind growth and predicted values
      return(pred.res)
    } else {                                         # skip if doesn't converge
      pred = data.frame(x=NA,y=NA) %>%
        nest()
      res <- data.frame(phi1=NA,phi2=NA,phi3=NA,flag='try-error',
                        stringsAsFactors = F)
      fit.stats <- as.data.frame(matrix(nrow = 1,ncol = 11,dimnames = list(1,fit.stat.cols))) # empty fit.stats
      pred.res <- bind_cols(pred,res,fit.stats)
      return(pred.res)
    } 
  } else {    
    pred = data.frame(x=NA,y=NA) %>%
      nest()
    res <- data.frame(phi1=NA,phi2=NA,phi3=NA,flag="few-time-points",
                      stringsAsFactors = F)
    fit.stats <- as.data.frame(matrix(nrow = 1,ncol = 11,
                                      dimnames = list(1,fit.stat.cols))) # empty fit.stats
    pred.res <- bind_cols(pred,res,fit.stats)
    return(pred.res)
  }
}
```

### Helper function to extract predictions for desired input 'x' values (e.g. days, thermal days).
```{r}
getPred = function(phi1, phi2, phi3){
  x <- seq(0,2500,by=50)  #construct a range of x values for thermal time
  y <- try(phi1/(1+exp(-phi2*(x - phi3))),silent=TRUE)  #predicted y values  <old: exp(-(phi2+phi3*x)))
  if(class(y)!="try-error"){ 
    pred <- data.frame(x,y)                                #create the prediction data frame
    return(pred)
  } else {
    pred <- data.frame(x=NA,y=NA)
    return(pred)
  }
}
```

## Now lets get started with our analysis by reading the data file. Enter `help(read.csv)` in the console to know about the *read.csv* function.
```{r read-data, echo=TRUE}
# read curated data UAV height data 
dat.ht <- read.csv("C:/Users/singhdj2/Documents/daljit/r_stuff/tutorials/data/18LDH-tutorial-log-regression_2019-06-19.csv",stringsAsFactors = F,header = T)
head(dat.ht)
```

## There are five columns in the data file. Column *condition* refers to the optimum and late sowing conditions; *plot_id* is unique plot_id for 1200 plots in each condition; *thermal.time* is the cumulative sum of the mean daily air temperature at each condition; DAS is Days after sowing (in Julian Day units); *phenotype_value* is the height value estimated from drone based images.

## Before moving forward with your analysis, it is always good to take a look at the data structure and column attributes . Here we will take a quick glimpse at our data file. 
```{r data-checks, echo=TRUE}
glimpse(dat.ht) 
# set the condition as 'factor' for plotting convenience
dat.ht <- as_tibble(dat.ht) %>%
  mutate(condition=factor(condition,levels=c("Optimum","Late")))
```


## Let us find out how many unique time-points are there per condition?
```{r calc-time-points, echo=TRUE}
dat.ht %>% 
  group_by(condition) %>% 
  summarize(time.points=length(unique(thermal.time)))
```


## In which condition the plants grow taller? Lets get mean height per time-point for each condition to find out.
```{r get-mean-height, echo=TRUE}
dat.ht %>% 
  group_by(condition,thermal.time) %>%
  summarise(ht.mean=mean(phenotype_value,na.rm=T))
```


## We can dig a little deeper to look at the raw height trends at both field conditions.
```{r plot-height-trends, warning=FALSE}
x.label = "Thermal Time (°Cd)"    
ggplot(data=dat.ht, aes(y = phenotype_value, 
                        x = thermal.time, 
                        group=interaction(thermal.time,condition), 
                        fill=condition)) +
  geom_boxplot(width=25,alpha=0.7,notch = TRUE, size=0.8,
               notchwidth = 0.5, outlier.alpha = 0.2) +
  labs(x=x.label,y="Height (cm)") +
  scale_y_continuous(breaks= seq(0,150,20)) +
  scale_x_continuous(breaks= seq(0,2400,200)) +
  theme_minimal() +
  theme(panel.border = element_rect(colour = "black", fill = NA, size = .5),
        axis.text.x = element_text(colour = "black", size = 11,angle = 0),
        axis.text.y = element_text(colour = "black", size = 11),
        plot.title = element_text(size = 14, face = "bold"),
        text = element_text(size=14,face="bold"),
        legend.position="top") +
  guides(fill=guide_legend(title="")) +
  annotate("rect", xmin=1650, xmax=2320, ymin=96, ymax=101, alpha=0.3, fill="red") +
  annotate("rect", xmin=1400, xmax=2100, ymin=90, ymax=95, alpha=0.3, fill="#0da3a7")
# We can also save the above plot by using ggsave function.
# create a file name character string
# fName = paste0('../data/18LDH_tutorial-height-trend_boxplot_',Sys.Date(),'.jpg')
# # save the plot on disk
# ggsave(fName, plot = last_plot(), device = "jpg", path = "output/",
#        scale = 1, width = 10, height = 6.5, units = c("in"), dpi=800)
```

## Question: By chance, did you notice different height trends at two conditions? Which condition has plants growing faster/slower?


## We have data collected at multuple time-points. This rich temporal information can be summarized into a few growth parameters that have a straightforward biological interpretation. To do so, we will deploy our logistic regression function (remember we defined logistic function at the beginning of this script!) for each plot and condition using group_by and do functions from tidyverse package. 
```{r run-logRegr, warning=FALSE, cache=TRUE}
dat.ht.growth <- dat.ht %>%
  group_by(condition,plot_id) %>%    
  do(runLogReg(dates = .$thermal.time, phenValue = .$phenotype_value)) %>%
  ungroup() 
head(dat.ht.growth)
```
## Note that the parameters phi1, phi2, phi3 correspond to the maximum plant height (cm), growth rate (cm/thermal time), thermal time when plant is halfway its maximum height (measured in degree Cd units), respectively.


## Now that we have succefully summarized our time-points into three biologically interpretable parameters, we can go ahead and take a look at the regression fit summaries to find out how accurately our model was able to fit the original data.
```{r fit-summaries, warning=FALSE}
gghistogram(dat.ht.growth, x = "RMSE",
   add = "mean", rug = T,
   color = "condition", fill = "condition",
   palette = c("#00AFBB", "#E7B800")) +
  rremove("legend") +
  xlab("Root Mean Sq Error") +
  ylab("Count")
```

## Keep in mind that the R-square statistic is a bit tricky for non-linear models. Therefore, Efron.r-sq is used as a workaround approximation for checking the model fits in non-linear models. We will plot Efron.r-sq to check the fit of our regression model. 

```{r fit-R2, warning=FALSE}
gghistogram(dat.ht.growth, x = "Efron.r.squared",
   add = "mean", rug = T,
   #color = "condition",
   fill = "#00AFBB",
   facet.by = "condition") +
  rremove("legend") +
  xlab("Effron's R-Square") +
  ylab("Count")
```

## It looks like our logistic model fit our data really well. Almost 99% of the data had a perfect fit. 

## Now we are quite sure that our data fits worked well. Now we can ask some biological questions, for example, does sowing date have any effect on plant growth? Lets dig further...

### Let us check mean values for each growth parameter.
```{r calc-mean-params, warning=FALSE}
growth.mean <- dat.ht.growth %>% 
  group_by(condition) %>% 
  summarize_at(vars(contains('phi')),funs(mean(.,na.rm = T)))
growth.mean
```
## Looks like the late sown plants are gowing much faster in short time! 

## Wait...but, are these mean values statistically significant? 

### We can create violin plots and also perform mean comparisons for phi1 (maximum plant height) in a single step.
```{r plotting-phi1, warning=FALSE}
# Specify the comparisons you want
my_comparisons <- list( c("Optimum", "Late") )
# plot violins
dat.ht.growth %>% as_tibble() %>%
  ggviolin(., x = "condition", y = "phi1", 
           fill = "condition",
           palette = c("#00AFBB", "#E7B800"), #"#FC4E07"),
           #facet.by = "trait_id",
           #scales = "free",
           add = "boxplot", add.params = list(fill = "white")) +
  stat_compare_means(comparisons = my_comparisons, label = "p.signif")+ #Add significance levels
  stat_compare_means(label.y = 150) +
  rremove("legend") +
  xlab("Condition") +
  ylab("Digital Plant Height (cm)")
```

## What do you see in these plots? Does this figure answer our question of sowing date effect on plant height?

## Violin plots for phi2 (height growth rate)
```{r plotting-phi2, warning=FALSE}
dat.ht.growth %>% as_tibble() %>%
  ggviolin(., x = "condition", y = "phi2", 
           fill = "condition",
           palette = c("#00AFBB", "#E7B800"), #"#FC4E07"),
           #facet.by = "trait_id",
           #scales = "free",
           add = "boxplot", add.params = list(fill = "white")) +
  stat_compare_means(comparisons = my_comparisons, label = "p.signif")+ #Add significance levels
  stat_compare_means(label.y = 0.006) +
  rremove("legend") +
  xlab("Condition") +
  ylab("Growth Rate (cm/°Cd)")
```

## Does the late planted crop grow faster? If so, any guess why?


### Violin plots for phi3 (time to reach half maximum height).
```{r plotting-phi3, warning=FALSE}
## phi3 - plot violins with threshold
dat.ht.growth %>% as_tibble() %>%
  ggviolin(., x = "condition", y = "phi3", 
           fill = "condition",
           palette = c("#00AFBB", "#E7B800"), #"#FC4E07"),
           #facet.by = "trait_id",
           #scales = "free",
           add = "boxplot", add.params = list(fill = "white")) +
  stat_compare_means(comparisons = my_comparisons, label = "p.signif")+ # Add significance levels
  stat_compare_means(label.y = 1400) +
  rremove("legend") +
  xlab("Condition") +
  ylab("°Cd to Reach Half-maximum Height)")
```

## This plot is also consistent with what we see in previous two plots, i.e., the late sown crop grows significantly faster and take less time to reach to its half maximum height.

# Concluding remarks
## Through this tutorial you have familiarized yourself with a number of programming and statistical concepts namely: R functions, conditional workflows (if, else), tidy-data, exploratory data analysis, graphics, non-linear regression modeling, mean comparisons etc. However, always remeber that the tutorials such this one can only provide a primer or starting point- the real learning actually comes from a goal-oriented program exercises. Just like any other skill, programming is a game of regular practice. I would suggest you to pick a topic of your choice/interest and start playing with some example datasets; or better yet, if you have your own data try that out first.  
## As always, if you have any questions or interesting ideas to share, please free to reach out to me or other members of the K-State PBG journal club. 

# Happy Coding!!!


```{r session-info}
sessionInfo()
```
